\name: A Non-Parametric, Hierarchical \\ Wavelength-Calibration Model for a Precision Spectrograph

### 2020-05-14 paper/paper.tex
Full Incorporation of Bedell Suggestions
Discussion of applying to other spectrographs, other possible implementations, other things...

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{Figures/methodCartoon.pdf}
\caption{}
\label{fig:testHists}
\end{figure*}

@@

In the implementation of \name\ presented here, the magnitudes of the principal components are interpolated with respect to time.  New magnitudes $a_n'k$ are found for the midpoint time of non-calibration exposures and used to construct the calibration state of the spectrograph for that exposure.  Using interpolated magnitudes, $a_{n;k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, \name\ can construct a set of calibration lines, $x(\lambda,m,n') $, for any exposure as prescribed in Equation \ref{eq:excl_wsol}.  \lz{The best interpolation scheme for \expres\' instrument drift is discussed below in Section \ref{sec:choice_avt}, but may vary for different instruments.}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/lineDensity.png}
\caption{\lz{(Figure I use for talks, maybe shouldn't be included here)}  Comparison of polynomial-fit and interpolated wavelength solutions using either just ThAr lines or LFC lines. Left: the difference between each wavelength solution evaluated at the position of the LFC lines and the assigned wavelength of that LFC line.  The right plot is simply a zoom in of the left plot.  Note the large deviation when trying to implement Excalibur with just ThAr lines, which exhibit much large separations than LFC lines.  The classic, polynomial  fit exhibits similar residuals regardless of whether the set of ThAr lines or LFC lines are used.  Excalibur using LFC lines exhibit the lowest residuals.}
\label{fig:waveResids}
\end{figure*}

The residuals of the classic, polynomial wavelength solution is shown in the left-most plot of Figure \ref{fig:resid2d}.  There is a lot of vertical structure and some hints of a periodic, diagonal structure as well.  \lz{(Possibly from detector effects?  We're double checking this now and thinking about approaching STA about it.  Debra was considering publishing the results from that discussion in this paper if it seems relevant.)}.

The residuals of the interpolation test for the same exposure is shown in the center plot of Figure \ref{fig:resid2d}.  There is no coherent structure here and smaller residuals.  We believe the flexibility of the interpolated model was able to account for high-order detector defects, which emerged as structure in the residuals of the classic, smooth, polynomial-driven wavelength solution.

@@

\section{Choice Choices} \label{sec:choices}
This implementation of \name\ presented here incorporated determining various global variables and methods that we believe are or are close to optimal for constructing a high-fidelity wavelength solution.  This section will describe each choice and the associated decision-making process.

@@

In deciding a K value, we ran denoising tests, as described in Section \ref{sec:tests}, for K values spanning from 2 to 512.  One would expect the returned wavelengths to get increasingly more accurate with larger K until components that represent only noise are incorporated.  We found that the returned wavelengths have the lowest residuals with a K value of 32 (see Figure \ref{fig:kvals}).  However, the improvement is slight with K values greater than six.  Comparisons of wavelengths returned by a K=6 model vs. a K=32 model show significant differences for less than 10, bluer lines.  A visual inspection of the principal components, such as the one displayed in Figure \ref{fig:pcLfc}, led us to believe most of the structure is being captured in the first six principal components.

\begin{figure}[t]
\centering
\includegraphics[width=.1\textwidth]{Figures/pcsLfc6.png}
\caption{Spread in residuals as a function of $K$.}
\label{fig:kvals}
\end{figure}

\subsection{Interpolation of $a_{nk}$ with Respect to Time}
\label{sec:choice_avt}
Figure \ref{fig:nightlyVariation} shows the amplitude of the first and second principal component with respect to time on the left.  Though there exists a complex overall shape to the amplitudes with respect to time, a clear linear trend exists within each night.  This is shown by the right plots in Figure  \ref{fig:nightlyVariation}.  As the beginning-of-night and end-of-night calibration sets always include LFC exposures, we use a simple linear interpolation to interpolate principal component amplitudes with respect to time.

\lz{Yeah, the right-figure is kind of a mess, isn't it?}

We differentiated eras of stability by where the principal component amplitudes showed large variation.  The analysis was done by eye here, though many algorithms for change-point detection exist in the literature \lz{(Am I going to have to cite things?)}.  We were very conservative with defining eras of stability to ensure that within each era, the calibration-space of the instrument is low-dimensional as assumed.

\mb{(Can we quantify or illustrate how much change is too much change, for use in discussion of other instruments?)}

\begin{figure*}[h!]
\centering
\includegraphics[width=0.56\textwidth]{Figures/pcA_lfc.png}
\includegraphics[width=0.42\textwidth]{Figures/pcAs_byDay.png}
\caption{Amplitude of the first two principal components show as a function of time (left) or fraction of a day (right).  The top plots shows the amplitudes for the first principal component while the bottom plot shows the amplitudes for the second principal component.  Points mark the times at which we have calibration data.  Lines show the result of a linear interpolation.  In the right plot, the principal component amplitudes have been artificially offset by the median amplitude for each day so that all days are roughly on the same scale.  Here, points are colored by the MJD of each exposure.}
\label{fig:nightlyVariation}
\end{figure*}

\lz{Add empirical determination of how often EXPRES needs to redo calibration exposures?}

@@

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{Figures/intpx_tests.png}
\caption{Results from different interpolation schemes over pixels.  ThAr lines, shown as pink points, are used to construct a wavelength solution that is then evaluated at each LFC line, shown as black lines.  Histograms of the residuals for each method is shown on the right.  Note: there is a blended ThAr line at approximately pixel 5300.}
\label{fig:xinterp}
\end{figure*}

@@

@@

\mb{Add discussion of applicability to non-EPRV instruments.  Use ThAr results to comment on instruments w/o very dense calibration sources.}
\lz{We can maybe use the temperature swing in September to discuss less-stable spectrographs?  But it will remain hard to properly characterize ``stability'', and I'm not sure how representative that is for other instruments.  I think I manually did it by warding against high-order changes in the principal component amplitudes?  But presumably there exists an instrument for which the whole ability to implement PCA won't be true.}

>>>>

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{Figures/methodCartoon.pdf}
\caption{A cartoon representation of the \name\ method, as described in Section \ref{sec:method}.  Variations in line position, amplitude of basis vectors, and interpolation greatly enlarged to show texture.  In step one, dimensionality reduction and denoising, the complete set of line positions for each exposure is analyzed to return a set of basis vectors, $g_n$, which span the calibration space of the spectrograph.  In step two, \name\ interpolates the amplitudes of each basis vector, $a_n,k$, to return the calibration state for a specific exposure, given as a set of de-noised lines.  The assigned wavelengths of these de-noised line positions are then interpolated onto other pixels in step three to construct a full wavelength model.}
\label{fig:testHists}
\end{figure*}

@@

In the implementation of \name\ presented here, the magnitudes of the principal components are interpolated with respect to time.  New magnitudes $a_n'k$ are found for the midpoint time of non-calibration exposures and used to construct the calibration state of the spectrograph for that exposure.  Using interpolated magnitudes, $a_{n;k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, \name\ can construct a set of calibration lines, $x(\lambda,m,n') $, for any exposure as prescribed in Equation \ref{eq:excl_wsol}.  \lz{The best interpolation scheme for \expres\ is discussed below in Section \ref{sec:choice_avt}, but may vary for different instruments.}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{Figures/lineDensity.png}
\caption{Comparison of polynomial-fit and interpolated wavelength solutions using either just ThAr lines or LFC lines.  For a subset of one order, the difference between the assigned and model wavelengths for each LFC line is plotted with respect to pixel.  The large residuals between two widely-spaced ThAr lines when implementing an interpolated wavelength solution with just ThAr lines (blue curve) extends out of frame.  The classic, polynomial fit exhibits similar residuals in both amplitude and shape regardless of whether the set of ThAr lines or LFC lines are used.  An interpolated wavelength solution using LFC lines (orange curve) exhibits the lowest residuals.}
\label{fig:waveResids}
\end{figure}

The residuals of the classic, polynomial wavelength solution is shown in the left-most plot of Figure \ref{fig:resid2d}.  There is a lot of vertical structure and some hints of a periodic, diagonal structure as well.  The residuals of the interpolation test for the same exposure is shown in the center plot of Figure \ref{fig:resid2d}.  There is no coherent structure here and smaller residuals.  We believe the flexibility of the interpolated model was able to account for high-order detector defects, which emerged as structure in the residuals of the classic, smooth, polynomial-driven wavelength solution.  In the \expres\ data, for example, the interpolated model accounted for \lz{detector effect STA will explain to us}.

@@

\section{Choose Your Own Implementation} \label{sec:choices}
We describe here only one implementation of \name.  Using PCA and an interpolated wavelength solution is a fairly trivial step towards a complete hierarchical and non-parametric wavelength model.  It is possible to upgrade both the denoising and wavelength solution steps to true models.

\lz{Hogg: please edit/add to these methods}

For the dimensionality reduction and denoising, the PCA could be replaced by a probabilistic PCA model or other probabilistic linear reduction methods.  It is also possible to move to non-linear reduction methods, like...?

The wavelength solution could also move past interpolation.  For example, a Gaussian process could be used that is constrained to ensure monotonicity.  Replacing each step with a model will allow for full, hierarchical Bayesian inference.  This means error from wavelength calibration could be marginalized out.  This is only worth doing, however, if wavelength calibration is a significant fraction of the error budget.

The implementation of \name\ presented here, using PCA for denoising and interpolating a wavelength solution, incorporated various global variables and methods that we believe are or are close to optimal for constructing a high-fidelity wavelength solution.  The following subsections will describe each choice and the associated decision-making process.

@@

In deciding a K value, we ran denoising tests, as described in Section \ref{sec:tests}, for K values spanning from 2 to 512.  One would expect the returned wavelengths to get increasingly more accurate with larger K until components that represent only noise are incorporated.  We found that the returned wavelengths plateau at a K value of 32 (see Figure \ref{fig:kvals}).  However, the improvement is slight with K values greater than six.  Comparisons of wavelengths returned by a K=6 model vs. a K=32 model show significant differences for less than 10 bluer lines, which are know to have greater error in their measured line positions.  A visual inspection of the principal components, such as the one displayed in Figure \ref{fig:pcLfc}, additionally show most of the structure is being captured in the first six principal components with later components dominated by noise.

\begin{figure}[t]
\centering
\includegraphics[width=.5\textwidth]{Figures/kvals_all.pdf}
\caption{Spread in residuals of returned wavelengths for different values of $K$.  There is a dotted, vertical line at $K=6$, our chosen $K$ value.  The dotted, horizontal line is the RMS for $K=6$.}
\label{fig:kvals}
\end{figure}

\subsection{Interpolation of Calibration State with Respect to Time}
\label{sec:choice_avt}
Figure \ref{fig:nightlyVariation} shows the amplitude of the first and second principal component with respect to time on the left.  Though there exists a complex overall shape to the amplitudes with respect to time, a clear linear trend exists within each night.  This is shown by the right plots in Figure  \ref{fig:nightlyVariation}.  As the beginning-of-night and end-of-night calibration sets always include LFC exposures, we use a simple linear interpolation to interpolate principal component amplitudes with respect to time.

The choice in interpolation method can help guide how many wavelength calibration images are truly needed.  It is unnecessary to take calibration images at times where similar information can be filled in by a well-chosen interpolation scheme.  For example, with the \expres\ data shown here, it is clear that nightly calibration images are needed, but for a linear trend, only two calibration images throughout the night are strictly required.

The amplitudes $a_{n,k}$ can also be interpolated with respect to any housekeeping data, not just time.  Best results should come from interpolating with respect to whichever data is most strongly correlated with the calibration state of the spectrograph.  For example, with \expres, which is passively temperature controlled, the returned amplitudes $a_{n,k}$ were extremely correlated with the enclosure temperature, suggesting it would also be possible to interpolate with respect to temperature.

Another pertinent example would be a spectrograph that is mounted on the telescope, and therefore moves with the telescope.  In this case, it may be important to interpolate at least in part with respect to the position of the telescope, which enables the resultant calibration to incorporate the gravitational loading experienced by the spectrograph.

\begin{figure*}[h!]
\centering
\includegraphics[width=0.56\textwidth]{Figures/pcA_lfc.png}
\includegraphics[width=0.42\textwidth]{Figures/pcAs_byDay.png}
\caption{Amplitude of the first two principal components show as a function of time (left) or fraction of a day (right) \lz{(Yeah, it's kind of a mess)}.  The top plots shows the amplitudes for the first principal component while the bottom plot shows the amplitudes for the second principal component.  Points mark the times at which we have calibration data.  Lines show the result of a linear interpolation.  In the right plot, the principal component amplitudes have been artificially offset by the median amplitude for each day so that all days are roughly on the same scale.  Here, points are colored by the MJD of each exposure.}
\label{fig:nightlyVariation}
\end{figure*}

@@

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{Figures/intpx_tests.png}
\caption{Residuals from different interpolation schemes over pixels.  ThAr lines, shown as black circles, are used to construct a wavelength solution that is then evaluated at each LFC line, shown as black lines.  Histograms of the residuals for each method is shown on the right.  Note: there is a blended ThAr line at approximately pixel 5300, the last ThAr line plotted.}
\label{fig:xinterp}
\end{figure*}

@@

\section{Application to Other Spectrographs}
We use \expres\ data here not because it is the optimal system for a method like \name, but because there is a strong need for good wavelength calibration in the EPRV community.  The applicability of \name\ to any one instrument is a detailed question of the kind variance experienced by the spectrograph and calibration sources available, but we hope to provide some approximate benchmarks and diagnostics.

To determine the needed line density, we systematically removed LFC lines from the \expres\ data and calculated the RMS of the returned wavelengths.  Figure \ref{fig:density} plots the increase in the scatter of the wavelengths returned by \name\ as a function of separation between lines.  However, the stability of the lines and their measured line centers quickly becomes the dominant error source in calibration lamps over the separation between lines.

In terms of stability, \name\ should be adaptable to a wide range of instrument designs.  In truth, the amplitude of the variation matters little as long as all axes of variation can be captured by PCA or another dimensionality-reducing model.  Indeed, larger variations from less stabilized instruments might even make it easier for the PCA to identify these axes.  Most physical systems should have a few dominant axis along which they vary, simply by what is possible with mechanical engineering.  With PCA, this can be tested by plotting the power of the basis vectors describing the calibration space, which should fall quickly after a few vectors.  Note, however, that this is only a measure of the variance in the PCA space, not an explicit test of the end-to-end variation in the resulting model.  This condition is therefore necessary, but not sufficient if implementing \name with PCA.

It could still be possible to run \name\ on a spectrograph that has a high-dimensional calibration space, meaning a large number of basis vectors is required to capture all the variance in the spectrograph.  In this regime, there is always the risk of over-fitting.  Regularizing the principal component amplitudes, for example insisting the amplitudes for higher principal components be smaller, can help to return reasonable models \citep{formanmackey2015} .  Within such a framework, \name\ may still deliver good results.
\lz{Hogg: My notes said you might want to add more things here?}

For the results presented here, the data was broken up into different eras of stability based on where the principal component amplitudes showed huge deviations.  This was done visually, though there exists many change-point detection algorithms that could be used.  There is a trade-off in including more exposures between introducing greater variation, but also more data to offer constraints that may be optimized.  Here, an era of stability was chosen  manually in order to focus on separating out time-domain intervals in which the data is relatively homogeneous, e.g. most exposures show the same lines.  Homogeneity is, of course, implicitly required when implementing PCA.

@@
